### 2010-06-30 article.tex
First stuff getting put in

\documentclass[11pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage{url}

\title{A Theoretical Approach to Virtual Memory}
\author{Peter Boothe\\
Manhattan College\\
\url{peter.boothe@manhattan.edu}}
\date{Submission to SODA 2011} % or maybe ALENEX2011

\newcommand{\heapsort}{{\sc HeapSort}}
\newcommand{\quicksort}{{\sc QuickSort}}
\newcommand{\bubblesort}{{\sc BubbleSort}}

\begin{document}
\maketitle

As memory requirements begin to outstrip available memory on a machine, the
operating system will allocate virtual memory to supplement the available
physical memory.  We present, argue for, and experimentally validate a simple
analysis technique which allows algorithms researchers to analyze the behavior
of an algorithm when using virtual memory.  In particular, the penalty for
fetching a page from virtual memory can be extreme, and can make
previously-small constants quite large. Thus, it would be of great use to be
able to easily estimate, for a given algorithm, the number of times the
algorithm will have to pause to retrieve virtual memory off the hard drive.

\section{The Method}

In the worst case, every memory access will have to go to the hard drive.
This, however, is not an informative worst-case analysis.  The worst-case page
fault predictions do not reflect the real-world behavior of virtual memory ---
they reflect a situation in which every cache access is a cache miss!  Instead,
our method is to simply count the number of non-adjacent memory and array
accesses.  Using this, we find that we also end up providing a quantitative
measure of ``cache friendliness'', which is a term often used, but seldomly
rigorously defined.  We will demonstrate our method using three classic sorting
techniques: \heapsort, \quicksort, and \bubblesort.

Our analysis of \heapsort is the easiest.  The \heapsort algorithm is simply 
\begin{verbatim}
heapsort(array, size)
    make-heap(array, size)
    for i = size-1 ... 0
        array[i] = remove-min(array, i+1)
    return array
\end{verbatim}

\end{document}

### 2010-07-01 sorts/heapsort.c
Adding in some data generation code

@@ -1,0 +1,24 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <bsd/stdlib.h>
+
+int cmp(const void *i, const void *j)
+{
+        if (*((int*)i) == *((int*)j)) return 0;
+        else return *((int*)i) < *((int*)j) ? 1 : -1;
+}
+
+#define Mb (1024*1024)
+
+int main(int argc, char **argv)
+{
+        int *array, i, SIZE;
+        SIZE = atoi(argv[1])*Mb/sizeof(int);
+
+        array = malloc(sizeof(int) * SIZE);
+        for (i = 0; i < SIZE; i++) array[i] = i*7 % SIZE;
+
+        heapsort(array, SIZE, sizeof(int), &cmp);
+
+        return 0;
+}

### 2010-07-01 sorts/memkiller.c

@@ -1,0 +1,18 @@
+#include <stdlib.h>
+#include <time.h>
+
+#define Mb (1024*1024)
+
+int main(int argc, char **argv)
+{
+        long long size;
+        char *mem;
+        size = atol(argv[1]) * Mb;
+
+        mem = (char *)malloc(size * sizeof(char));
+
+        srandom(time(NULL));
+        while (1) {
+                mem[random() % size] = (char)(random() % 256);
+        }
+}

### 2010-07-01 sorts/quicksort.c

@@ -1,0 +1,23 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+int cmp(const void *i, const void *j)
+{
+        if (*((int*)i) == *((int*)j)) return 0;
+        else return *((int*)i) < *((int*)j) ? 1 : -1;
+}
+
+#define Mb (1024*1024)
+
+int main(int argc, char **argv)
+{
+        int *array, i, SIZE;
+        SIZE = atoi(argv[1])*Mb/sizeof(int);
+
+        array = malloc(sizeof(int) * SIZE);
+        for (i = 0; i < SIZE; i++) array[i] = i*7 % SIZE;
+
+        qsort(array, SIZE, sizeof(int), &cmp);
+
+        return 0;
+}

### 2010-07-01 article.tex
It's beginning to look like a paper...

@@
 \usepackage{url}
+\usepackage{salgorithm}
 
-\title{A Theoretical Approach to Virtual Memory}
-\author{Peter Boothe\\
+\title{A Simple and Useful Model of Virtual Memory Performance}
+\author{Peter Boothe and Gregory Rae\\
 Manhattan College\\
-\url{peter.boothe@manhattan.edu}}
+\url{peter.boothe@manhattan.edu}
+\\
+\url{greg.rae@gmail.com}
+}
 \date{Submission to SODA 2011} % or maybe ALENEX2011
@@
 \newcommand{\quicksort}{{\sc QuickSort}}
-\newcommand{\bubblesort}{{\sc BubbleSort}}
+\newcommand{\mergesort}{{\sc MergeSort}}
+\newcommand{\NAA}{\textrm{NAA}}
 
@@
 physical memory.  We present, argue for, and experimentally validate a simple
-analysis technique which allows algorithms researchers to analyze the behavior
-of an algorithm when using virtual memory.  In particular, the penalty for
-fetching a page from virtual memory can be extreme, and can make
+model and analysis technique which allows algorithm developers to analyze the
+behavior of an algorithm when virtual memory is used.  In particular, the
+penalty for a page fault in virtual memory can be extreme, and can make
 previously-small constants quite large. Thus, it would be of great use to be
 able to easily estimate, for a given algorithm, the number of times the
-algorithm will have to pause to retrieve virtual memory off the hard drive.
+algorithm will have to wait for the operating system to retrieve virtual memory
+pages off the hard drive.
 
@@
 fault predictions do not reflect the real-world behavior of virtual memory ---
-they reflect a situation in which every cache access is a cache miss!  Instead,
-our method is to simply count the number of non-adjacent memory and array
-accesses.  Using this, we find that we also end up providing a quantitative
-measure of ``cache friendliness'', which is a term often used, but seldomly
-rigorously defined.  We will demonstrate our method using three classic sorting
-techniques: \heapsort, \quicksort, and \bubblesort.
+they reflect a situation in which every cache access is a cache miss! Our
+model predicts that all non-adjacent memory accesses have a uniform non-zero probability of causing a page fault.
 
-Our analysis of \heapsort is the easiest.  The \heapsort algorithm is simply 
-\begin{verbatim}
-heapsort(array, size)
-    make-heap(array, size)
-    for i = size-1 ... 0
-        array[i] = remove-min(array, i+1)
-    return array
-\end{verbatim}
+Our method of estimating the number of page faults is therefore to count the
+number of non-adjacent memory accesses.  Using this, we find that we also end
+up providing a quantitative measure of ``cache friendliness'', which is a term
+often used, but seldom rigorously defined.  We demonstrate our method
+using three classic sorting techniques: \heapsort, \quicksort, and \mergesort.
+
+\subsection{\heapsort\ Page Fault Analysis}
+Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
+take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
+supporting functions, in Figure~\ref{fig:hsort}.
+
+\begin{figure}
+\begin{algorithm}
+\Algorithm{HeapSort}$(array, size)$\+\\
+    Build-Max-Heap$(array, size)$\\
+    \For $i \gets size$ \DownTo\ $2$\+\\
+        exchange $array[1] \leftrightarrow array[i]$\\
+        $size \gets size - 1$\\
+        Max-Heapify$(array, 1, size)$\-\\
+    \Return $array$\-\\
+\\
+\Algorithm{Build-Max-Heap}$(array, size)$\+\\
+    \For $i \gets \lfloor \frac{size}{2} \rfloor$ \DownTo\ $1$\+\\
+        Max-Heapify$(array, i, size)$\-\-\\
+\\
+\Algorithm{Max-Heapify}$(array, i, size)$\+\\
+    $l \gets 2*i$\\
+    $r \gets 2*i + 1$\\
+    $largest \gets i$\\
+    \If $l < size$ and $array[l] > array[largest]$\+\\
+            $largest \gets l$\-\\
+    \If $r < size$ and $array[r] > array[largest]$\+\\
+            $largest \gets r$\-\\
+    \If $largest \neq i$\+\\
+        exchange $array[i] \leftrightarrow array[largest]$\\
+        Max-Heapify$(array, largest, size)$
+\end{algorithm}
+
+\caption{\heapsort\ from Cormen {\it et al.}\cite{clrs}.}
+\label{fig:hsort}
+\end{figure}
+
+In the pseudocode of \heapsort, we note that, with the exception of the very
+top of the heap, at no point do we ever access memory adjacent to the most
+recently used memory location.  Therefore, we predict that the number of page
+faults will be proportional to the number of memory accesses, which in the case
+of \heapsort\ is proportional to its runtime of $O(n \log n)$.
+
+\subsection{\quicksort\ Page Fault Analysis}
+The \quicksort\ algorithm will be dealt with later.
+
+\subsection{\mergesort\ Page Fault Analysis}
+The \mergesort\ algorithm, which we also
+take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
+supporting functions, in Figure~\ref{fig:msort}.
+In our analysis, the only consideration is the number and location of memory accesses, and memory accesses only occur in the ``merge'' portion of the algorithm.
+
+\begin{figure}
+\begin{algorithm}
+\Algorithm{MergeSort}$(array, p, r)$\+\\
+    \If $p < r$\+\\
+        $q \gets \lfloor\frac{p+r}{2}\rfloor$\\
+        MergeSort$(array, p, q)$\\
+        MergeSort$(array, q+1, r)$\\
+        Merge$(array, p, q, r)$\-\-\\
+\\
+\Algorithm{Merge}$(array, p, q, r)$\+\\
+    $n_1 \gets q - p + 1$\\
+    $n_2 \gets r - q$\\
+    create arrays $L[1\ldots n_1+1]$ and $R[1\ldots n_2+1]$\\
+    \For $i \gets 1 \ldots n_1$\+\\
+        $L[i] \gets array[p+i-1]$\-\\
+    \For $j \gets 1 \ldots n_2$\+\\
+        $R[j] \gets array[q+j]$\-\\
+    $L[n_1+1] \gets \infty$, $R[n_2+1] \gets \infty$\\
+    $i \gets 1$, $j \gets 1$\\
+    \For $k \gets p \ldots r$\+\\
+        \If $L[i] \le R[j]$\+\\
+            $array[k] \gets L[i]$\\
+            $i \gets i+1$\-\\
+        \Else\+\\
+            $array[k] \gets R[j]$\\
+            $j \gets j+1$
+\end{algorithm}
+
+\caption{\mergesort\ from Cormen {\it et al.}\cite{clrs}.}
+\label{fig:msort}
+\end{figure}
+
+In merging, the first step is to access the beginning and middle of the
+passed-in array (two non-adjacent accesses). Then, two extra arrays are
+created, the contents of the passed-in array are copied over to the new arrays
+(many accesses, only the first access to each array is non-adjacent).  Then, in
+the merge step, the contents of the two new arrays are merged back into the
+main array (many accesses, but only the first access to each array is
+non-adjacent).
+
+In total, each call to the merge function will require five non-adjacent memory
+accesses.  Therefore, over the whole of the algorithm, we find that the number
+of non-adjacent accesses ($\NAA(n)$) is $$\NAA(n) = 5 + 2  \NAA(\frac{n}{2})$$
+Immediately, by the master theorem, we can see that $\NAA(n) \in O(n)$, and
+that we therefore expect the number of page faults to grow linearly with the
+size of the input to \mergesort.
+
+\section{Experimental Validation of Our Model}
+\subsection{Experimental Setup}
+\subsection{\heapsort}
+\subsection{\quicksort}
+\subsection{\mergesort}
+
+\section{Conclusion}
+
+\bibliographystyle{abbrv}
+\bibliography{bibliography}
 

### 2010-07-01 sorts/heapsort.c
Little changes to fix overflow errors

@@ -12,10 +12,11 @@ int cmp(const void *i, const void *j)
 
 int main(int argc, char **argv)
 {
-        int *array, i, SIZE;
-        SIZE = atoi(argv[1])*Mb/sizeof(int);
+        int *array, i;
+        long SIZE;
+        SIZE = atol(argv[1])/sizeof(int)*Mb;
 
-        array = malloc(sizeof(int) * SIZE);
+        array = malloc(SIZE*sizeof(int));
         for (i = 0; i < SIZE; i++) array[i] = i*7 % SIZE;
 
         heapsort(array, SIZE, sizeof(int), &cmp);

### 2010-07-02 sorts/heapsort.c
Moving towards a better analysis technique

@@ -12,14 +12,20 @@ int cmp(const void *i, const void *j)
 
 int main(int argc, char **argv)
 {
-        int *array, i;
+        int *array, i, repeats;
         long SIZE;
         SIZE = atol(argv[1])/sizeof(int)*Mb;
+        repeats = atoi(argv[2]);
 
         array = malloc(SIZE*sizeof(int));
-        for (i = 0; i < SIZE; i++) array[i] = i*7 % SIZE;
+        
+        while (repeats--> 0) { 
+                for (i = 0; i < SIZE; i++) array[i] = i*7 % SIZE;
 
-        heapsort(array, SIZE, sizeof(int), &cmp);
+                heapsort(array, SIZE, sizeof(int), &cmp);
+        }
+
+        free(array);
 
         return 0;
 }

### 2010-08-18 article.tex
Hooray!  Starting to make an article of it.

@@
 \usepackage{salgorithm}
+\usepackage{graphicx} \graphicspath{{sorts/}}
 
 \title{A Simple and Useful Model of Virtual Memory Performance}
-\author{Peter Boothe and Gregory Rae\\
+\author{Peter Boothe\\
 Manhattan College\\
 \url{peter.boothe@manhattan.edu}
-\\
-\url{greg.rae@gmail.com}
 }
-\date{Submission to SODA 2011} % or maybe ALENEX2011
+\date{Submission to ALENEX2011}
 
@@
 take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
-supporting functions, in Figure~\ref{fig:hsort}.
+supporting functions, in Figure~\ref{fig:hsort}.  The implementation we tested
+is from the BSD standard library, and libbsd on Linux.
 
@@
 
-\subsection{\quicksort\ Page Fault Analysis}
-The \quicksort\ algorithm will be dealt with later.
+\subsection{\mergesort\ Page Fault Analysis}
 
-\subsection{\mergesort\ Page Fault Analysis}
-The \mergesort\ algorithm, which we also
-take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
-supporting functions, in Figure~\ref{fig:msort}.
-In our analysis, the only consideration is the number and location of memory accesses, and memory accesses only occur in the ``merge'' portion of the algorithm.
+The \mergesort\ algorithm we take from Cormen {\it et al.}\cite{clrs} with the
+memory-efficiency improvement that we only copy out the first half of the
+array.  Our implementation, in its entirety, may be found in
+Figure~\ref{fig:msort}.  In our analysis, the only consideration is the number
+and location of memory accesses, and memory accesses only occur in the
+``merge'' portion of the algorithm.
 
 \begin{figure}
-\begin{algorithm}
-\Algorithm{MergeSort}$(array, p, r)$\+\\
-    \If $p < r$\+\\
-        $q \gets \lfloor\frac{p+r}{2}\rfloor$\\
-        MergeSort$(array, p, q)$\\
-        MergeSort$(array, q+1, r)$\\
-        Merge$(array, p, q, r)$\-\-\\
-\\
-\Algorithm{Merge}$(array, p, q, r)$\+\\
-    $n_1 \gets q - p + 1$\\
-    $n_2 \gets r - q$\\
-    create arrays $L[1\ldots n_1+1]$ and $R[1\ldots n_2+1]$\\
-    \For $i \gets 1 \ldots n_1$\+\\
-        $L[i] \gets array[p+i-1]$\-\\
-    \For $j \gets 1 \ldots n_2$\+\\
-        $R[j] \gets array[q+j]$\-\\
-    $L[n_1+1] \gets \infty$, $R[n_2+1] \gets \infty$\\
-    $i \gets 1$, $j \gets 1$\\
-    \For $k \gets p \ldots r$\+\\
-        \If $L[i] \le R[j]$\+\\
-            $array[k] \gets L[i]$\\
-            $i \gets i+1$\-\\
-        \Else\+\\
-            $array[k] \gets R[j]$\\
-            $j \gets j+1$
-\end{algorithm}
+\begin{verbatim}
+void mergesort(int array[], long lo, long hi)
+{
+    int *spare;
+    int ssize;
+    long sindex, hindex;
+    long mid;
 
-\caption{\mergesort\ from Cormen {\it et al.}\cite{clrs}.}
+    // Base case
+    if (lo+1 >= hi) return;
+
+    // Recurse on each half
+    mid = (lo + hi) / 2;  
+    mergesort(array, lo, mid);
+    mergesort(array, mid, hi);
+
+    // Merge
+    ssize = mid-lo; // Only copy out the first half
+    spare = (int *)malloc(ssize * sizeof(int));
+    memcpy(spare, &(array[lo]), ssize * sizeof(int));
+
+    sindex = 0;
+    hindex = mid;
+    while (sindex < ssize && hindex < hi) {
+        if (spare[sindex] < array[hindex])
+            array[lo++] = spare[sindex++];
+        else
+            array[lo++] = array[hindex++];
+    }
+
+    while (sindex < ssize) array[lo++] = spare[sindex++];
+
+    free(spare);
+}
+\end{verbatim}
+
+\caption{Our \mergesort\ implementation in its entirety.}
 \label{fig:msort}
@@
 
-In merging, the first step is to access the beginning and middle of the
-passed-in array (two non-adjacent accesses). Then, two extra arrays are
-created, the contents of the passed-in array are copied over to the new arrays
-(many accesses, only the first access to each array is non-adjacent).  Then, in
-the merge step, the contents of the two new arrays are merged back into the
-main array (many accesses, but only the first access to each array is
-non-adjacent).
+In merging, the first step is to allocate a new array of size equal to half of
+the passed-in array.  Then, the first half of the passed-in array is copied to
+the new array.  The copying is done by accessing consecutive elements of each
+array, and so once the beginning of each array has been accessed, all subsequent memory accesses are adjacent.  Therefore, the initial step has two non-adjacent accesses.
 
-In total, each call to the merge function will require five non-adjacent memory
-accesses.  Therefore, over the whole of the algorithm, we find that the number
-of non-adjacent accesses ($\NAA(n)$) is $$\NAA(n) = 5 + 2  \NAA(\frac{n}{2})$$
-Immediately, by the master theorem, we can see that $\NAA(n) \in O(n)$, and
-that we therefore expect the number of page faults to grow linearly with the
+Subsequently, we treat the latter half of the passed-in array as one queue, and the newly-allocated array as the second queue.
+We then merge by repeatedly removing the smaller item from the front of the two queues and overwriting an element of the passed-in array.
+Every examination requires a memory access, but in all cases except the initial examination of the fronts of the queues, all memory accesses occur adjacent to previous memory accesses.
+Therefore, to merge the newly-allocated array and the passed-in array requires two non-adjacent memory accesses.
+In total, each call to our \mergesort function can require four non-adjacent memory accesses.
+
+Over the whole of the algorithm, we find that the number
+of non-adjacent accesses for \mergesort on an array of size $n$ ($\NAA(n)$) is $$\NAA(n) = 4 + 2\NAA(\frac{n}{2})$$
+By the master theorem, we can see that $\NAA(n) \in \Theta(n)$, and
+we therefore expect the number of page faults to grow linearly with the
 size of the input to \mergesort.
 
-\section{Experimental Validation of Our Model}
+\section{Experimental Validation of Our Model's Conclusions}
+
+An experimental model can not be verified to be correct, it can only be shown
+that the available tests do not prove the model wrong.  Here, we construct a
+set of tests which repeatedly fail to disprove our model.
+
 \subsection{Experimental Setup}
+
+To analyze the number of major page faults in an algorithm, we must run our
+algorithm in low-memory conditions so that the operating system is forced to
+allocate and use virtual memory for our program.  Unfortunately, accessing swap
+is incredibly slow --- a major part of the motivation for our study --- which
+means that under low-memory conditions our tests may require 100x more time to
+finish.
+
+We reconcile the opposing forces by running our tests in a virtual machine via
+the Qemu emulator.  The machine which is running the emulator was given 4
+gigabytes of memory.  Of those 4 Gb, half of it was allocated to be a RAM disk.
+The swap partition of the emulated machine was a file stored on that RAM disk.
+This meant that the page faults of the emulated machine would still be
+recorded, but that the emulated machine would not suffer a performance hit, as
+the emulated swap was backed by an extremely fast RAM disk!  We recommend this
+technique for anyone who would like to count page faults but does not want to
+waste a lot of time.
+
+Interestingly, the RAM disk turned out to be helpful, but not essential, as the
+natural caching behaviour of the host operating system meant that even when the
+emulated swap was backed by a physical hard drive, it was often cached in the
+memory of the host operating system.
+
 \subsection{\heapsort}
-\subsection{\quicksort}
+
 \subsection{\mergesort}
+
+\subsection{Number of Page Faults Compared}
+
+\begin{figure}
+\input{sorts/memsizes.tex}
+\caption{Number of major page faults for \mergesort\ and \heapsort\ at a variety of memory sizes.  Note that at almost all memory sizes, \mergesort\ is better.}
+\end{figure}
+
+\begin{figure}
+\input{sorts/ratios.tex}
+\caption{Number major page faults for \heapsort\ divided by the number of major page faults for \mergesort\ at a variety of memory sizes.
+The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model!}
+\end{figure}
 

### 2010-08-18 sorts/memsizes.sh

@@ -1,20 +1,33 @@
 #!/bin/bash
 
+TERM=png
+OUTPUT=
+if [ -n "$1" ]
+then
+        TERM=$1
+        OUTPUT=$2
+fi
+
 cat - <<EOF
-set term png
+set term ${TERM}
 set title "Problem size vs Page Faults\\nEach line holds available memory constant"
 set yrange [0:1000000]
 set xrange [0:1100]
 set style line 1 linewidth 2 pt 4 linetype 3 ps 2
 set style line 2 linewidth 2 linetype 10 ps 2
 set xlabel "Problem size in megabytes"
-set ylabel "# of major page faults"
+set ylabel "Number of major page faults"
 set border 0
 set key below
 set grid
 set mxtics 
 set mytics 
 EOF
+
+if [ -n "$OUTPUT" ]
+then
+        echo set output \"${OUTPUT}\"
+fi
 
 ( echo plot
 KEYED=

### 2010-08-18 sorts/ratios.sh

@@ -1,2 +1,41 @@
 #!/bin/bash
 
+TERM=png
+OUTPUT=
+if [ -n "$1" ]
+then
+        TERM=$1
+        OUTPUT=$2
+fi
+
+cat - <<EOF
+set term ${TERM}
+set title "Heapsort is bad and gets worse as problems grow\nEach line holds available memory constant"
+
+set xlabel "Problem size in megabytes"
+set ylabel "Ratio of Heapsort page faults to Mergesort page faults"
+set border 0
+set key below
+set grid
+set mxtics 
+set mytics 
+set yrange [0:20]
+set key off
+EOF
+
+if [ -n "$OUTPUT" ]
+then
+        echo set output \"${OUTPUT}\"
+fi
+
+for i in `seq 1 30`
+do
+        echo set style line $i linewidth 2 ps 2
+done
+
+c=1
+for i in ratios/*.dat
+do
+        echo \\\"$i\\\" using 1:\\\(\$3/\$2\\\) with linespoints ls $c,
+        c=$[ $c + 1 ]
+done | xargs echo plot | sed -e 's/,$//'

### 2010-08-22 article.tex
Now works with the required ACM style.

@@
-\documentclass[11pt]{article}
-\usepackage[letterpaper]{geometry}
+\documentclass{acmtrans2m}
 \usepackage{url}
@@
 \usepackage{graphicx} \graphicspath{{sorts/}}
+\usepackage{fancyvrb}
 
-\title{A Simple and Useful Model of Virtual Memory Performance}
-\author{Peter Boothe\\
-Manhattan College\\
-\url{peter.boothe@manhattan.edu}
+%\acmVolume{V} 
+%\acmNumber{N} 
+%\acmYear{YY} 
+%\acmMonth{M} 
+\markboth{Peter Boothe}{A Simple, Useful Model of Virtual Memory} 
+\title{A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory}
+\author{PETER BOOTHE\\
+Manhattan College
 }
-\date{Submission to ALENEX2011}
+\begin{abstract} 
+We define a simple model of virtual memory in an effort to predict the number of page faults that an algorithm will occur over the course of its lifetime.
+We then use that model to analyze how many page faults we would expect to see with two of the best-known and most well-studied sorting algorithms.
+Then, we run multiple experiments in an effort to show that the predictions of our model match up with reality.
+
+\end{abstract}
+
+\category{F.2.0}{Analysis of Algorithms and Problem Complexity}{General} 
+\category{D.4.2}{Operating Systems}{Storage Management} [Virtual Memory]
+\category{D.4.8}{Operating Systems}{Performance} [Modeling and Prediction]
+\terms{Algorithms, Measurement, Experimentation} 
+\keywords{Virtual Memory, Algorithm Analysis}
 
@@
 \begin{document} 
+\begin{bottomstuff} 
+This research was supported by a summer grant from Manhattan College.\\
+Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463\\
+\end{bottomstuff} 
 \maketitle
@@
 operating system will allocate virtual memory to supplement the available
-physical memory.  We present, argue for, and experimentally validate a simple
-model and analysis technique which allows algorithm developers to analyze the
-behavior of an algorithm when virtual memory is used.  In particular, the
+physical memory.  There are several theoretical approaches to dealing with
+virtual memory, but none of these methods are easily used after an algorithm is
+already designed or implemented.  Even worse, some of the methods require
+knowledge of available resources and the manual management of a process'
+virtual memory.  In this paper we introduce a model of virtual memory that has
+the benefits of being both simple and useful.  As designed, it is also slightly
+incorrect, but it seems to work in practice nonetheless.  This new method
+allows an algorithm designer or implementer to quickly derive how many page
+faults their algorithm will occur on an input of size $n$, assuming the input
+will not fit in available physical memory.  We describe the model
+(Section~\ref{model}), carry out two sample analyses using the model
+(Section~\ref{analysis}), and then experimentally validate the conclusions of
+our analysis (Section~\ref{experiment}).
+
+ The
 penalty for a page fault in virtual memory can be extreme, and can make
@@
 pages off the hard drive.
+On a modern system, the CPU can perform 2 billion operations a second, but the access time to retrieve a page of virtual memory from a consumer hard drive can take as multiple milliseconds.
+Even if we assume ultra-fast hard drives, a lower bound of 1 millisecond seems more than resonable, which mean that the process is stalled, waiting for memory, for more than 2 million wasted CPU operations.
+If each tick of the CPU clock took a second, then virtual memory would impose a penalty of more than 23 days!
+Clearly, it would be good to be able to asymptotically compare the behavior of different algorithms in the presence of virtual memory, so as to minimize the number of page faults.
 
-\section{The Method}
+\section{The Model}
+\label{model}
 
@@
 often used, but seldom rigorously defined.  We demonstrate our method
-using three classic sorting techniques: \heapsort, \quicksort, and \mergesort.
+using two classic sorting techniques: \heapsort\ and \mergesort.
+
+Before we continue, however, we should address one obvious fault of our model --- it is not completely correct!
+In particular, if we wrote a program that traversed an array that was twice the size of available memory, our model predicts that we will see at most 1 page fault, when in actuality there will be at least $\frac{n}{2}$.
+This objection is correct, as far as it goes, and means that our model can be ``gamed'' and should only serve as a lower-bound measure of page faults if exactness is required.
+However, the extreme simplicity of our model has a lot of appeal, so we will conduct our experiments and see if the predictions of our (slightly wrong) model are useful and accurate.
+
+\section{Using the Model to Analyze Existing Algorithms}
+\label{analysis}
+
+In this section, we use our new model to analyze already-developed algorithms.
+In particular, we focus on \heapsort\ and \mergesort, because it is a folk theorem that \heapsort\ is not ``cache friendly'', while \mergesort\ is the opposite.
+If our analysis can concretize and lend rigor to this folk theorem, then our model might be useful in other situations as well.
 
@@
 Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
-take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
+take from \citeN{clrs}, is defined, with all of its
 supporting functions, in Figure~\ref{fig:hsort}.  The implementation we tested
-is from the BSD standard library, and libbsd on Linux.
+is from the BSD standard library, which is available as {\tt libbsd} on Linux.
 
-\begin{figure}
+\begin{narrowfig}{1.5in}
 \begin{algorithm}
@@
 
-\caption{\heapsort\ from Cormen {\it et al.}\cite{clrs}.}
+\caption{\heapsort\ pseudo code.}
 \label{fig:hsort}
-\end{figure}
+\end{narrowfig}
 
@@
 
-The \mergesort\ algorithm we take from Cormen {\it et al.}\cite{clrs} with the
-memory-efficiency improvement that we only copy out the first half of the
-array.  Our implementation, in its entirety, may be found in
-Figure~\ref{fig:msort}.  In our analysis, the only consideration is the number
-and location of memory accesses, and memory accesses only occur in the
-``merge'' portion of the algorithm.
+The \mergesort\ algorithm we adapt from \citeN{clrs} with the
+memory-efficiency improvement that, in the merge step, we only copy out the
+first half of the array, and merge the second half in-place.  Our
+implementation may be found in Figure~\ref{fig:msort}.  In
+our analysis of page faults, the only consideration is the number and location
+of memory accesses, and memory accesses only occur in the ``merge'' portion of
+the algorithm.
 
 \begin{figure}
-\begin{verbatim}
+{\footnotesize
+\begin{Verbatim}[numbers=left,numbersep=3pt,xleftmargin=20pt]
 void mergesort(int array[], long lo, long hi)
@@
     while (sindex < ssize && hindex < hi) {
-        if (spare[sindex] < array[hindex])
+        if (spare[sindex] <= array[hindex])
             array[lo++] = spare[sindex++];
@@
 }
-\end{verbatim}
+\end{Verbatim}
+}
 
-\caption{Our \mergesort\ implementation in its entirety.}
+\caption{Our \mergesort\ implementation, in C, in its entirety.}
 \label{fig:msort}
@@
 Therefore, to merge the newly-allocated array and the passed-in array requires two non-adjacent memory accesses.
-In total, each call to our \mergesort function can require four non-adjacent memory accesses.
+In total, each call to our \mergesort\ function can require four non-adjacent memory accesses.
 
 Over the whole of the algorithm, we find that the number
-of non-adjacent accesses for \mergesort on an array of size $n$ ($\NAA(n)$) is $$\NAA(n) = 4 + 2\NAA(\frac{n}{2})$$
+of non-adjacent accesses for \mergesort\ on an array of size $n$ (which we will call $\NAA(n)$) is $$\NAA(n) = 4 + 2\NAA\left(\frac{n}{2}\right)$$
 By the master theorem, we can see that $\NAA(n) \in \Theta(n)$, and
@@
 
-\section{Experimental Validation of Our Model's Conclusions}
+\section{Experimental Validation of Model Conclusions}
+\label{experiment}
 
-An experimental model can not be verified to be correct, it can only be shown
+A model can not be experimentally verified to be correct, it can only be shown
 that the available tests do not prove the model wrong.  Here, we construct a
@@
 is incredibly slow --- a major part of the motivation for our study --- which
-means that under low-memory conditions our tests may require 100x more time to
-finish.
+means that under low-memory conditions our tests may require 1000 times more
+time to finish than they would if we weren't using swap.
 
-We reconcile the opposing forces by running our tests in a virtual machine via
-the Qemu emulator.  The machine which is running the emulator was given 4
-gigabytes of memory.  Of those 4 Gb, half of it was allocated to be a RAM disk.
-The swap partition of the emulated machine was a file stored on that RAM disk.
-This meant that the page faults of the emulated machine would still be
-recorded, but that the emulated machine would not suffer a performance hit, as
-the emulated swap was backed by an extremely fast RAM disk!  We recommend this
-technique for anyone who would like to count page faults but does not want to
-waste a lot of time.
+We reconcile the opposing forces of our desire for speed and our desire to use
+swap by running our tests in a virtual machine via the Qemu emulator.  The
+machine running the emulator had 4 gigabytes of memory.  Of those 4 gigabytes,
+half was allocated as a RAM disk.  The swap partition of the emulated machine
+was a file stored on that RAM disk.  This meant that the page faults of the
+emulated machine would still be recorded, but that the emulated machine would
+not suffer a performance hit, as the emulated swap was backed by a (very fast)
+RAM disk!  This technique is recommended for anyone who would like to count
+page faults but does not want to waste a lot of time.
 
-Interestingly, the RAM disk turned out to be helpful, but not essential, as the
+Interestingly, the RAM disk turned out to be helpful, but not essential. The
 natural caching behaviour of the host operating system meant that even when the
 emulated swap was backed by a physical hard drive, it was often cached in the
-memory of the host operating system.
+memory of the host operating system due to the automatic buffering of hard drive reads and writes.
 
-\subsection{\heapsort}
+\subsection{\heapsort\ and \mergesort\ Implementations and Test Harness}
 
-\subsection{\mergesort}
+For \heapsort\ we used the implementation found in the BSD standard library, which is available as {\tt libbsd} on Linux.
+For \mergesort\ we wrote our own implementation, which can be found in Figure~\ref{fig:msort}.
+The test harness that we used was exactly the same for both algorithms, and can be seen in Figure~\ref{fig:harness}.
+
+\begin{narrowfig}{3in}
+\begin{Verbatim}[numbers=left,numbersep=3pt,xleftmargin=20pt]
+#include <stdio.h>
+#include <stdlib.h>
+
+/// SORTING ALGORITHM DEFINITION GOES HERE
+
+#define Mb (1024*1024)
+
+int main(int argc, char **argv)
+{
+        int *array, i;
+        long SIZE;
+        SIZE = atol(argv[1])*Mb/sizeof(int);
+	printf("Allocated %ld ints\n", SIZE);
+
+        array = malloc(SIZE*sizeof(int));
+        fprintf(stderr, "%p\n", array);
+
+        for (i = 0; i < SIZE; i++) {
+            array[i] = i*7 % SIZE;
+        }
+
+        /// CALL TO SORTING ALGORITHM GOES HERE
+
+        free(array);
+
+        return 0;
+}
+\end{Verbatim}
+\caption{Our test harness for both algorithms}
+\label{fig:harness}
+\end{narrowfig}
 
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{acmtrans}
 \bibliography{bibliography}
 
+\begin{received} 
+Received Month Year; revised Month Year; accepted Month Year 
+\end{received}
 \end{document}

### 2010-08-23 article.tex
Fixed lines

@@
 \documentclass{acmtrans2m}
+
+\def\endbottomstuff{\end{figure}\gdef\permission{}}
+
+\def\runningfoot{\def\@runningfoot{}}
+\def\firstfoot{\def\@firstfoot{}}
+
 \usepackage{url}
@@
 \begin{abstract} 
-We define a simple model of virtual memory in an effort to predict the number of page faults that an algorithm will occur over the course of its lifetime.
-We then use that model to analyze how many page faults we would expect to see with two of the best-known and most well-studied sorting algorithms.
-Then, we run multiple experiments in an effort to show that the predictions of our model match up with reality.
-
+We define a simple model of virtual memory in an effort to predict the number
+of page faults that an algorithm will occur over the course of its lifetime.
+We then use that model to analyze how many page faults we would expect to see
+with two of the best-known and most well-studied sorting algorithms.  Then, we
+run multiple experiments in an effort to show that the predictions of our model
+match up with reality.
 \end{abstract}
@@
 This research was supported by a summer grant from Manhattan College.\\
-Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463\\
+Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463
 \end{bottomstuff} 
@@
 
- The
-penalty for a page fault in virtual memory can be extreme, and can make
+The penalty for a page fault in virtual memory can be extreme, and can make
 previously-small constants quite large. Thus, it would be of great use to be
@@
 algorithm will have to wait for the operating system to retrieve virtual memory
-pages off the hard drive.
-On a modern system, the CPU can perform 2 billion operations a second, but the access time to retrieve a page of virtual memory from a consumer hard drive can take as multiple milliseconds.
-Even if we assume ultra-fast hard drives, a lower bound of 1 millisecond seems more than resonable, which mean that the process is stalled, waiting for memory, for more than 2 million wasted CPU operations.
-If each tick of the CPU clock took a second, then virtual memory would impose a penalty of more than 23 days!
-Clearly, it would be good to be able to asymptotically compare the behavior of different algorithms in the presence of virtual memory, so as to minimize the number of page faults.
+pages off the hard drive.  On a modern system, the CPU can perform 2 billion
+operations a second, but the access time to retrieve a page of virtual memory
+from a consumer hard drive can take as multiple milliseconds.  Even if we
+assume ultra-fast hard drives, a lower bound of 1 millisecond seems more than
+resonable, which mean that the process is stalled, waiting for memory, for more
+than 2 million wasted CPU operations.  If each tick of the CPU clock took a
+second, then virtual memory would impose a penalty of more than 23 days!
+Clearly, it would be good to be able to asymptotically compare the behavior of
+different algorithms in the presence of virtual memory, so as to minimize the
+number of page faults.
 
@@
 fault predictions do not reflect the real-world behavior of virtual memory ---
-they reflect a situation in which every cache access is a cache miss! Our
-model predicts that all non-adjacent memory accesses have a uniform non-zero probability of causing a page fault.
+they reflect a situation in which every cache access is a cache miss. Our model
+predicts that all non-adjacent memory accesses have a uniform non-zero
+probability of causing a page fault.
 
@@
 up providing a quantitative measure of ``cache friendliness'', which is a term
-often used, but seldom rigorously defined.  We demonstrate our method
-using two classic sorting techniques: \heapsort\ and \mergesort.
+often used, but seldom rigorously defined.  We demonstrate our method using two
+classic sorting techniques: \heapsort\ and \mergesort.
 
-Before we continue, however, we should address one obvious fault of our model --- it is not completely correct!
-In particular, if we wrote a program that traversed an array that was twice the size of available memory, our model predicts that we will see at most 1 page fault, when in actuality there will be at least $\frac{n}{2}$.
-This objection is correct, as far as it goes, and means that our model can be ``gamed'' and should only serve as a lower-bound measure of page faults if exactness is required.
-However, the extreme simplicity of our model has a lot of appeal, so we will conduct our experiments and see if the predictions of our (slightly wrong) model are useful and accurate.
+Before we continue, however, we should address one obvious fault of our model
+--- it is not completely correct!  In particular, if we wrote a program that
+traversed an array that was twice the size of available memory, our model
+predicts that we will see at most 1 page fault, when in actuality there will be
+at least $\frac{n}{2}$.  This objection is correct, as far as it goes, and
+means that our model can be ``gamed'' and should only serve as a lower-bound
+measure of page faults if exactness is required.  However, the extreme
+simplicity of our model has a lot of appeal, so we will conduct our experiments
+and see if the predictions of our (slightly wrong) model are nonetheless useful
+and accurate.
 
@@
 In this section, we use our new model to analyze already-developed algorithms.
-In particular, we focus on \heapsort\ and \mergesort, because it is a folk theorem that \heapsort\ is not ``cache friendly'', while \mergesort\ is the opposite.
-If our analysis can concretize and lend rigor to this folk theorem, then our model might be useful in other situations as well.
+In particular, we focus on \heapsort\ and \mergesort, because it is a folk
+theorem that \heapsort\ is not ``cache friendly'', while \mergesort\ is the
+opposite.  If our analysis can concretize and lend rigor to this folk theorem,
+then our model might be useful in other situations as well.
 
 \subsection{\heapsort\ Page Fault Analysis}
+
 Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
-take from \citeN{clrs}, is defined, with all of its
-supporting functions, in Figure~\ref{fig:hsort}.  The implementation we tested
-is from the BSD standard library, which is available as {\tt libbsd} on Linux.
+take from \citeN{clrs}, is defined, with all of its supporting functions, in
+Figure~\ref{fig:hsort}.  The implementation we tested is from the BSD standard
+library, which is available as {\tt libbsd} on Linux.
 
@@
 
-The \mergesort\ algorithm we adapt from \citeN{clrs} with the
-memory-efficiency improvement that, in the merge step, we only copy out the
-first half of the array, and merge the second half in-place.  Our
-implementation may be found in Figure~\ref{fig:msort}.  In
-our analysis of page faults, the only consideration is the number and location
-of memory accesses, and memory accesses only occur in the ``merge'' portion of
-the algorithm.
+The \mergesort\ algorithm we adapt from \citeN{clrs} with the memory-efficiency
+improvement that, in the merge step, we only copy out the first half of the
+array, and merge the second half in-place.  Our implementation may be found in
+Figure~\ref{fig:msort}.  In our analysis of page faults, the only consideration
+is the number and location of memory accesses, and memory accesses only occur
+in the ``merge'' portion of the algorithm.
 
@@
 the new array.  The copying is done by accessing consecutive elements of each
-array, and so once the beginning of each array has been accessed, all subsequent memory accesses are adjacent.  Therefore, the initial step has two non-adjacent accesses.
+array, and so once the beginning of each array has been accessed, all
+subsequent memory accesses are adjacent.  Therefore, the initial step has two
+non-adjacent accesses.
 
-Subsequently, we treat the latter half of the passed-in array as one queue, and the newly-allocated array as the second queue.
-We then merge by repeatedly removing the smaller item from the front of the two queues and overwriting an element of the passed-in array.
-Every examination requires a memory access, but in all cases except the initial examination of the fronts of the queues, all memory accesses occur adjacent to previous memory accesses.
-Therefore, to merge the newly-allocated array and the passed-in array requires two non-adjacent memory accesses.
-In total, each call to our \mergesort\ function can require four non-adjacent memory accesses.
+Subsequently, we treat the latter half of the passed-in array as one queue, and
+the newly-allocated array as the second queue.  We then merge by repeatedly
+removing the smaller item from the front of the two queues and overwriting an
+element of the passed-in array.  Every examination requires a memory access,
+but in all cases except the initial examination of the fronts of the queues,
+all memory accesses occur adjacent to previous memory accesses.  Therefore, to
+merge the newly-allocated array and the passed-in array requires two
+non-adjacent memory accesses.  In total, each call to our \mergesort\ function
+can require four non-adjacent memory accesses.
 
-Over the whole of the algorithm, we find that the number
-of non-adjacent accesses for \mergesort\ on an array of size $n$ (which we will call $\NAA(n)$) is $$\NAA(n) = 4 + 2\NAA\left(\frac{n}{2}\right)$$
-By the master theorem, we can see that $\NAA(n) \in \Theta(n)$, and
-we therefore expect the number of page faults to grow linearly with the
-size of the input to \mergesort.
+Over the whole of the algorithm, we find that the number of non-adjacent
+accesses for \mergesort\ on an array of size $n$ (which we will call $\NAA(n)$)
+is $$\NAA(n) = 4 + 2\NAA\left(\frac{n}{2}\right)$$ By the master theorem, we
+can see that $\NAA(n) \in \Theta(n)$, and we therefore expect the number of
+page faults to grow linearly with the size of the input to \mergesort.
 
@@
 not suffer a performance hit, as the emulated swap was backed by a (very fast)
-RAM disk!  This technique is recommended for anyone who would like to count
+RAM disk.  This technique is recommended for anyone who would like to count
 page faults but does not want to waste a lot of time.
@@
 emulated swap was backed by a physical hard drive, it was often cached in the
-memory of the host operating system due to the automatic buffering of hard drive reads and writes.
+memory of the host operating system due to the automatic buffering of hard
+drive reads and writes.
 
@@
 
-For \heapsort\ we used the implementation found in the BSD standard library, which is available as {\tt libbsd} on Linux.
-For \mergesort\ we wrote our own implementation, which can be found in Figure~\ref{fig:msort}.
-The test harness that we used was exactly the same for both algorithms, and can be seen in Figure~\ref{fig:harness}.
+For \heapsort\ we used the implementation found in the BSD standard library,
+which is available as {\tt libbsd} on Linux.  For \mergesort\ we wrote our own
+implementation, which can be found in Figure~\ref{fig:msort}.  The test harness
+that we used was exactly the same for both algorithms, and can be seen in
+Figure~\ref{fig:harness}.
 
@@
 \caption{Number major page faults for \heapsort\ divided by the number of major page faults for \mergesort\ at a variety of memory sizes.
-The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model!}
+The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model.}
 \end{figure}
@@
 \begin{received} 
-Received Month Year; revised Month Year; accepted Month Year 
+Submission to ALENEX 2011
 \end{received}

### 2010-10-06 article.tex
Final touches.

@@
 We define a simple model of virtual memory in an effort to predict the number
-of page faults that an algorithm will occur over the course of its lifetime.
+of page faults that an algorithm will incur over the course of its lifetime.
 We then use that model to analyze how many page faults we would expect to see
@@
 allows an algorithm designer or implementer to quickly derive how many page
-faults their algorithm will occur on an input of size $n$, assuming the input
+faults their algorithm will incur on an input of size $n$, assuming the input
 will not fit in available physical memory.  We describe the model
@@
 operations a second, but the access time to retrieve a page of virtual memory
-from a consumer hard drive can take as multiple milliseconds.  Even if we
-assume ultra-fast hard drives, a lower bound of 1 millisecond seems more than
+from a consumer hard drive can be multiple milliseconds.  Even if we assume
+ultra-fast hard drives, a lower bound of 1 millisecond seems more than
 resonable, which mean that the process is stalled, waiting for memory, for more
 than 2 million wasted CPU operations.  If each tick of the CPU clock took a
-second, then virtual memory would impose a penalty of more than 23 days!
-Clearly, it would be good to be able to asymptotically compare the behavior of
-different algorithms in the presence of virtual memory, so as to minimize the
-number of page faults.
+second, then a virtual memory page fault would impose a penalty of more than 23
+days!  Clearly, it would be good to be able to asymptotically compare the
+behavior of different algorithms in the presence of virtual memory, so as to
+minimize the number of page faults.
+
+Prior work in this area is either sparse, or mind-blowingly large, depending on
+whether one counts the development of cache-oblivious algorithms\cite{cacheo},
+external memory sorting, and the optimization of sorting
+algorithms\cite{314324,radix} as related work.  Our purpose here, however, is a
+bit different.  Instead of designing an algorithm for low-memory conditions, we
+seek to take existing algorithms and be able to quickly analyze their
+performance when memory is scarce.
 
@@
 they reflect a situation in which every cache access is a cache miss. Our model
-predicts that all non-adjacent memory accesses have a uniform non-zero
+avoids the worst case, and instead predicts that all non-adjacent memory
+accesses have a uniform non-zero probability of causing a page fault.  Also, we
+make the simplifying assumption that function calls completely clear the cache,
+so all initial memory accesses after a function call have the same uniform
 probability of causing a page fault.
@@
 up providing a quantitative measure of ``cache friendliness'', which is a term
-often used, but seldom rigorously defined.  We demonstrate our method using two
-classic sorting techniques: \heapsort\ and \mergesort.
+often used, but seldom rigorously defined.  We demonstrate our method by
+comparing two classic sorting techniques: \heapsort\ and \mergesort.
 
-Before we continue, however, we should address one obvious fault of our model
---- it is not completely correct!  In particular, if we wrote a program that
-traversed an array that was twice the size of available memory, our model
-predicts that we will see at most 1 page fault, when in actuality there will be
-at least $\frac{n}{2}$.  This objection is correct, as far as it goes, and
-means that our model can be ``gamed'' and should only serve as a lower-bound
-measure of page faults if exactness is required.  However, the extreme
-simplicity of our model has a lot of appeal, so we will conduct our experiments
-and see if the predictions of our (slightly wrong) model are nonetheless useful
-and accurate.
+Before we continue, however, we should address one obvious fault of our
+model --- it is not completely correct!  In particular, if given a program that
+traversed an array twice the size of available memory, our model predicts that
+we will see at most 1 page fault, when in actuality there will be at least
+$O(\frac{n}{2})$ page faults.  This objection is correct, as far as it goes,
+and means that our model can be ``gamed'' and should only serve as a
+lower-bound measure of page faults if exactness is required.  However, the
+extreme simplicity of our model has great appeal, so we will conduct
+experiments and see if the predictions of our (admittedly slightly wrong) model
+are nonetheless useful and accurate.
 
@@
 In particular, we focus on \heapsort\ and \mergesort, because it is a folk
-theorem that \heapsort\ is not ``cache friendly'', while \mergesort\ is the
-opposite.  If our analysis can concretize and lend rigor to this folk theorem,
-then our model might be useful in other situations as well.
+theorem that \heapsort\ is not ``cache friendly'' and that \mergesort\ is
+``cache friendly''.  In this context, we take the term ``cache friendly'' to
+mean that we would expect the number of page faults from \heapsort\ to be
+higher than the number of page faults from \mergesort. If our analysis can
+concretize and lend rigor to this folk theorem, then our model might be useful
+in other situations as well.
 
@@
 that the available tests do not prove the model wrong.  Here, we construct a
-set of tests which repeatedly fail to disprove our model.
-
-\subsection{Experimental Setup}
-
-To analyze the number of major page faults in an algorithm, we must run our
-algorithm in low-memory conditions so that the operating system is forced to
-allocate and use virtual memory for our program.  Unfortunately, accessing swap
-is incredibly slow --- a major part of the motivation for our study --- which
-means that under low-memory conditions our tests may require 1000 times more
-time to finish than they would if we weren't using swap.
-
-We reconcile the opposing forces of our desire for speed and our desire to use
-swap by running our tests in a virtual machine via the Qemu emulator.  The
-machine running the emulator had 4 gigabytes of memory.  Of those 4 gigabytes,
-half was allocated as a RAM disk.  The swap partition of the emulated machine
-was a file stored on that RAM disk.  This meant that the page faults of the
-emulated machine would still be recorded, but that the emulated machine would
-not suffer a performance hit, as the emulated swap was backed by a (very fast)
-RAM disk.  This technique is recommended for anyone who would like to count
-page faults but does not want to waste a lot of time.
-
-Interestingly, the RAM disk turned out to be helpful, but not essential. The
-natural caching behaviour of the host operating system meant that even when the
-emulated swap was backed by a physical hard drive, it was often cached in the
-memory of the host operating system due to the automatic buffering of hard
-drive reads and writes.
-
-\subsection{\heapsort\ and \mergesort\ Implementations and Test Harness}
-
-For \heapsort\ we used the implementation found in the BSD standard library,
-which is available as {\tt libbsd} on Linux.  For \mergesort\ we wrote our own
-implementation, which can be found in Figure~\ref{fig:msort}.  The test harness
-that we used was exactly the same for both algorithms, and can be seen in
-Figure~\ref{fig:harness}.
+set of tests in an attempt to disprove our model.
 
@@
 \end{narrowfig}
+\subsection{Experimental Setup}
+
+To analyze the number of major page faults in an algorithm, we must run our
+algorithm in low-memory conditions so that the operating system is forced to
+allocate and use virtual memory for our program.  Unfortunately, accessing swap
+is incredibly slow --- a major part of the motivation for our study --- which
+means that under low-memory conditions our tests may require 1000000 times more
+time to finish than they would if we weren't using swap.
+
+We reconcile the opposing forces of our desire for speed and our desire to use
+swap by running our tests in a virtual machine via the Qemu emulator.  The
+machine running the emulator had 4 gigabytes of memory.  Of those 4 gigabytes,
+half was allocated as a RAM disk.  The swap partition of the emulated machine
+was a file stored on that RAM disk.  This meant that the page faults of the
+emulated machine would still be recorded, but that the emulated machine would
+not suffer a performance hit, as the emulated swap was backed by a (very fast)
+RAM disk.  This technique is recommended for anyone who would like to count
+page faults but does not want to waste a lot of time to do so.
+
+Interestingly, the RAM disk turned out to be helpful, but not essential. The
+natural caching behaviour of the host operating system meant that even when the
+emulated swap was backed by a file on a physical hard drive, it was often
+cached in the memory of the host operating system due to the host operating
+system's automatic buffering of hard drive reads and writes.
+
+\subsection{Implementations and Test Harness}
+
+For \heapsort\ we used the implementation found in the BSD standard library,
+which is available as {\tt libbsd} on Linux.  For \mergesort\ we wrote our own
+implementation, which can be found in Figure~\ref{fig:msort}.  The test harness
+that we used was exactly the same for both algorithms, and can be seen in
+Figure~\ref{fig:harness}.  To test our algorithms, we declared and allocated a
+single large array, and then we filled that array with many increasing
+sequences.
 
 \subsection{Number of Page Faults Compared}
+
+We compare the page faults of \heapsort\ and \mergesort\ for a wide range of memory sizes and problem sizes.
+We find that, for problems which do not fit into available memory, \heapsort\ incurs more page faults than \mergesort.
+Holding available memory constant, in Figure~\ref{fig:sizes} we see that \mergesort\ almost always has fewer page faults than \heapsort.
+We grew available memory in the virtual machine, in increments of 64 megabytes, from a low of 64 megabytes to a high of 512 megabytes.
+Each line in the figure represents the behavior with a fixed amount of available memory.
 
@@
 \caption{Number of major page faults for \mergesort\ and \heapsort\ at a variety of memory sizes.  Note that at almost all memory sizes, \mergesort\ is better.}
+\label{fig:sizes}
 \end{figure}
+
+As the ratio of problem size to available memory grows, so does the advantage
+of \mergesort\ over \heapsort.  In Figure~\ref{fig:ratio}, we can see the
+number of page faults of \heapsort\ divided by the number of page faults of
+\mergesort\ for a variety of problem sizes.  The ratio, as predicted, rises as
+the amount of memory pressure rises.  As problem sizes increase, \heapsort\
+performs increasingly worse compared to \mergesort. This experimental result is
+precisely in line with the results of our model which suggest that the number of page faults for \heapsort\ is asymptotically greater than the number of page faults in \mergesort\ ($O(n \log n)$ instead of $O(n)$).
 
@@
 The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model.}
+\label{fig:ratio}
 \end{figure}
 
+Because our experimental conclusions are in line with the conclusions of our
+model, we must consider this to be evidence in favor of the model's utility.
+Our experiments have failed to disprove our model, which means that it still
+may be correct enough for our purposes.
+
 \section{Conclusion}
+
+We have described a new model of virtual memory which has the virtue of extreme
+simplicity.
+In this model, every memory access which is not adjacent to a previous memory access has a uniform probability of incurring a virtual memory page fault.
+Furthermore, all function calls are considered to wipe out our caches, and so all memory accesses after a function call, even if adjacent to memory accessed before the function call, have that same uniform probability of causing a page fault.
+This model is, as stated, extremely simple (and a bit simplistic).
+
+In exchange for this simplicity, we had to sacrifice formal correctness.  Using
+experimental techniques, we validated the utility of our method, and showed
+that it does appear to be useful, and at least model the cases of \heapsort\
+and \mergesort\ well.  Much like a buggy program that still remains useful for
+some tasks, we find that our slightly-incorrect analysis technique can, with
+relative ease, provide useful estimates regarding the asymptotic behavior of
+algorithms in the presence of virtual memory.
+
+Future work in this domain would involve testing more algorithms under adverse
+memory conditions to see whether our model accurately reflects the behavior of
+more algorithms than just the two tested here.  Also, a deeper understanding of
+why our incorrect model gives such good and useful results (and when it can be
+counted upon to do so) would be of great interest.
 

### 2010-10-06 article.tex
References and cleanup

@@
 behavior of different algorithms in the presence of virtual memory, so as to
-minimize the number of page faults.
+minimize the number of page faults --- either asymptotically or just as a leading constant\cite{queue}.
 
@@
 
+\appendix
+
+\section{Code availability}
+
+All code written, all experimental data collected, and even this article may be found online at \url{http://github.com/pboothe/vmanalysis}
+
 \begin{received} 

### 2010-10-06 article.tex
11 pt font.  Looks uglier, conforms to guidelines.

@@
-\documentclass{acmtrans2m}
-
-\def\endbottomstuff{\end{figure}\gdef\permission{}}
-
-\def\runningfoot{\def\@runningfoot{}}
-\def\firstfoot{\def\@firstfoot{}}
+\documentclass[11pt]{article}
 
@@
 \usepackage{fancyvrb}
+\usepackage[letterpaper]{geometry}
 
@@
 %\acmMonth{M} 
+%
+%\category{F.2.0}{Analysis of Algorithms and Problem Complexity}{General} 
+%\category{D.4.2}{Operating Systems}{Storage Management} [Virtual Memory]
+%\category{D.4.8}{Operating Systems}{Performance} [Modeling and Prediction]
+%\terms{Algorithms, Measurement, Experimentation} 
+%\keywords{Virtual Memory, Algorithm Analysis} 
 \markboth{Peter Boothe}{A Simple, Useful Model of Virtual Memory} 
 \title{A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory}
-\author{PETER BOOTHE\\
+\author{Peter Boothe\thanks{This research was supported by a summer grant from Manhattan College.
+Author's Address: \texttt{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463
+}\\
 Manhattan College
 }
+
+\newcommand{\heapsort}{{\sc HeapSort}}
+\newcommand{\quicksort}{{\sc QuickSort}}
+\newcommand{\mergesort}{{\sc MergeSort}}
+\newcommand{\NAA}{\textrm{NAA}}
+
+\begin{document}
+
+\maketitle
+
 \begin{abstract} 
@@
 \end{abstract}
-
-\category{F.2.0}{Analysis of Algorithms and Problem Complexity}{General} 
-\category{D.4.2}{Operating Systems}{Storage Management} [Virtual Memory]
-\category{D.4.8}{Operating Systems}{Performance} [Modeling and Prediction]
-\terms{Algorithms, Measurement, Experimentation} 
-\keywords{Virtual Memory, Algorithm Analysis}
-
-\newcommand{\heapsort}{{\sc HeapSort}}
-\newcommand{\quicksort}{{\sc QuickSort}}
-\newcommand{\mergesort}{{\sc MergeSort}}
-\newcommand{\NAA}{\textrm{NAA}}
-
-\begin{document} 
-\begin{bottomstuff} 
-This research was supported by a summer grant from Manhattan College.\\
-Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463
-\end{bottomstuff} 
-\maketitle
 
@@
 Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
-take from \citeN{clrs}, is defined, with all of its supporting functions, in
+take from \cite{clrs}, is defined, with all of its supporting functions, in
 Figure~\ref{fig:hsort}.  The implementation we tested is from the BSD standard
@@
 
-\begin{narrowfig}{1.5in}
+\begin{figure}
 \begin{algorithm}
@@
 \label{fig:hsort}
-\end{narrowfig}
+\end{figure}
 
@@
 
-The \mergesort\ algorithm we adapt from \citeN{clrs} with the memory-efficiency
+The \mergesort\ algorithm we adapt from \cite{clrs} with the memory-efficiency
 improvement that, in the merge step, we only copy out the first half of the
@@
 
-\begin{narrowfig}{3in}
+\begin{figure}
 \begin{Verbatim}[numbers=left,numbersep=3pt,xleftmargin=20pt]
@@
 \label{fig:harness}
-\end{narrowfig}
+\end{figure}
 \subsection{Experimental Setup}
@@
 
-\bibliographystyle{acmtrans}
+\bibliographystyle{abbrv}
 \bibliography{bibliography}
@@
 
-\begin{received} 
+%\begin{received} 
 Submission to ALENEX 2011
-\end{received}
+%\end{received}
 \end{document}
